---
title: "TCGA Lung Cancer: LUAD vs LUSC — Molecular Subtypes, Immune Landscape & Clinical Outcomes"
author: "GoFigr Webinar"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    fig-width: 10
    fig-height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  dpi = 150
)
```

# Setup

```{r libraries}
library(gofigR)
library(SummarizedExperiment)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(pheatmap)
library(uwot)
library(survival)
library(survminer)
library(ggrepel)
library(patchwork)
```

## Enabling GoFigr

GoFigr is a figure management platform that automatically captures, versions, and indexes
every figure you produce. Once enabled, every plot rendered during this session is
published to GoFigr — no extra calls needed. You can browse, search, and share your
figures at [app.gofigr.io](https://app.gofigr.io).

To enable GoFigr, we call `gofigR::enable()`. All arguments are optional — called with
no arguments, `enable()` will read credentials from `~/.gofigr`, infer the workspace
from your config, and derive the analysis name from the document filename. We list them
here explicitly to show the key parameters that define a GoFigr session:

- **`workspace_name`** (default: your configured default workspace): The workspace that
  will contain this analysis. Workspaces are top-level organizational units — think of
  them like folders or projects. If a workspace with this name doesn't exist yet, GoFigr
  creates it automatically (controlled by `create_workspace`, which defaults to `TRUE`).
- **`analysis_name`** (default: inferred from the document filename): The analysis within
  the workspace. An analysis groups all figures from a single document or experiment.
  Again, created automatically if it doesn't exist.
- **`watermark`** (default: `QR_WATERMARK`): Controls what gets overlaid on each figure.
  `QR_WATERMARK` adds a QR code and link that points back to the figure on GoFigr, so
  anyone viewing the plot (in a paper, a slide deck, or a colleague's screen) can scan it
  to find the source data, code, and full revision history. Other options are
  `LINK_WATERMARK` (text link only) and `NO_WATERMARK`.

Behind the scenes, `enable()` reads your API key from `~/.gofigr` (generated once via
`gfconfig()`), authenticates with the GoFigr server, and installs hooks that intercept
`plot()` and `print()` calls to capture figures transparently. It also records the
execution context — which `.qmd` file produced each figure, which code chunk, and the
full `sessionInfo()` — so every figure is traceable back to the exact code that created it.

```{r gofigr}
gofigR::enable(
  workspace_name = "GoFigr Webinar (April 8, 2026)",
  analysis_name = "TCGA Lung Cancer Analysis (LUAD + LUSC)",
  watermark = QR_WATERMARK,
  auto_assign = TRUE,
  url="https://app-dev.gofigr.io"
)
```

# Data Loading & Preprocessing

We work with two TCGA lung cancer cohorts: LUAD (Lung Adenocarcinoma) and LUSC (Lung
Squamous Cell Carcinoma) — the two most common subtypes of non-small cell lung cancer
(NSCLC). Together they represent the most extensively characterized lung cancer datasets
available. The data was downloaded using `TCGAbiolinks` and saved as a Bioconductor
`SummarizedExperiment` — a container that bundles the expression matrix, gene annotations
(`rowData`), and clinical metadata (`colData`) into a single object.

The RDS file contains RNA-Seq TPM (transcripts per million) values, already subset to
~250 curated cancer genes plus the top 1000 most highly expressed genes (for unsupervised
analyses). The curated genes span driver mutations (EGFR, KRAS, ALK, etc.), immune
checkpoint and microenvironment markers, major signaling pathways (RTK/RAS, PI3K/AKT,
Wnt, Hippo, Notch), DNA damage response, EMT, metabolism, and lung-specific
differentiation markers.

We apply a log2(TPM + 1) transformation, which is standard for RNA-Seq expression
analysis: it stabilizes variance across the dynamic range and makes fold-changes
symmetric (a 2-fold increase and 2-fold decrease are equidistant from zero).

```{r load-data}
# Download the pre-built RDS if it doesn't already exist locally
data_dir <- file.path(getwd(), "data")
rds_path <- file.path(data_dir, "TCGA-LUNG_TPM_SE_subset.rds")

if (!file.exists(rds_path)) {
  dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
  message("Downloading TCGA data...")
  download.file(
    "https://cdn.gofigr.io/data/TCGA-LUNG_TPM_SE_subset.rds",
    rds_path,
    mode = "wb"
  )
}

gofigR::sync_file(rds_path)

se <- readRDS(rds_path)

# Extract expression matrix (TPM from unstranded counts) and log-transform
expr_raw <- assay(se, "tpm_unstrand")
expr_log <- log2(expr_raw + 1)

# Map to gene symbols, appending suffixes to any duplicates
gene_names <- rowData(se)$gene_name
rownames(expr_log) <- make.unique(gene_names)

expr <- expr_log

# Clinical metadata
clinical <- as.data.frame(colData(se))

cat(sprintf("Samples: %d | Genes: %d\n", ncol(expr), nrow(expr)))
cat(sprintf("Tumor: %d | Normal: %d\n",
            sum(clinical$sample_type == "Primary Tumor", na.rm = TRUE),
            sum(clinical$sample_type == "Solid Tissue Normal", na.rm = TRUE)))
```

TCGA includes both primary tumor and adjacent normal tissue samples. We restrict to
primary tumors — adjacent normals have fundamentally different biology and would dominate
any unsupervised analysis.

```{r filter-tumor}
tumor_idx <- clinical$sample_type == "Primary Tumor"
expr_tumor <- expr[, tumor_idx]
clin_tumor <- clinical[tumor_idx, ]

cat(sprintf("Tumor samples for analysis: %d\n", ncol(expr_tumor)))
```

# 1. Molecular Subtype Discovery

Our first goal is to see how LUAD and LUSC tumors relate at the transcriptomic level, and
whether finer-grained subtypes emerge within each histology. We start with PCA and UMAP
for visualization, then apply k-means clustering to identify discrete subtypes.

## PCA

Principal Component Analysis projects the high-dimensional expression data down to a
few axes that capture the most variance. We drop any genes with zero variance before
fitting, and let `prcomp` center the data (mean = 0 per gene). Since LUAD vs LUSC is a
major biological effect, we skip z-scaling so that genes with larger dynamic range can
contribute proportionally.

```{r pca}
gene_var <- apply(expr_tumor, 1, var, na.rm = TRUE)
expr_filt <- expr_tumor[gene_var > 0, ]

pca <- prcomp(t(expr_filt), center = TRUE, scale. = FALSE)
pct_var <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

pca_df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  PC3 = pca$x[, 3],
  stage = clin_tumor$ajcc_pathologic_stage,
  cohort = clin_tumor$cohort,
  gender = clin_tumor$gender,
  vital_status = clin_tumor$vital_status
)
```

We color by cancer type (LUAD vs LUSC) to see whether the dominant axis of variation
reflects histological identity — which we expect, given the profound biological
differences between adenocarcinoma and squamous cell carcinoma.

```{r pca-plot}
p <- ggplot(pca_df, aes(x = PC1, y = PC2, color = cohort)) +
  geom_point(alpha = 0.6, size = 2) +
  labs(
    x = sprintf("PC1 (%.1f%%)", pct_var[1]),
    y = sprintf("PC2 (%.1f%%)", pct_var[2]),
    title = "PCA of TCGA lung tumors (LUAD + LUSC)",
    subtitle = "Colored by cancer type",
    color = "Cohort"
  )
p %>% publish
```

## Top PCA loadings

The PCA rotation matrix tells us which genes contribute most to each principal component.
Since PC1 separates LUAD from LUSC, the top loadings should reflect the known biology of
these histologies — e.g. squamous differentiation markers (keratins, TP63, SOX2) vs
adenocarcinoma markers (NKX2-1/TTF-1, surfactant proteins, NAPSA).

```{r pca-loadings}
loadings <- as.data.frame(pca$rotation[, 1:3])
loadings$gene <- rownames(loadings)

top_pc1 <- loadings |>
  arrange(desc(abs(PC1))) |>
  head(15) |>
  mutate(gene = reorder(gene, abs(PC1)))

publish({ggplot(top_pc1, aes(x = PC1, y = gene)) +
  geom_col(aes(fill = PC1 > 0), show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "#c0392b", "FALSE" = "#2980b9")) +
  labs(
    title = "Top 15 genes driving PC1",
    x = "Loading on PC1",
    y = NULL
  )})
```

## LUAD vs LUSC: top differentially expressed genes

Since the dominant PCA axis separates cancer types, let's directly quantify the expression
differences. We compute log2 fold-change (LUSC vs LUAD) and a Welch t-test p-value for
each gene, then visualize the results as a volcano plot. This validates the PCA loadings
and highlights the biological basis of the separation: squamous markers (keratins, TP63,
SOX2) should be upregulated in LUSC, while adenocarcinoma markers (surfactant proteins,
NKX2-1, NAPSA) should be upregulated in LUAD.

```{r de-luad-lusc}
is_luad <- clin_tumor$cohort == "LUAD"
is_lusc <- clin_tumor$cohort == "LUSC"

de_results <- data.frame(
  gene = rownames(expr_filt),
  log2fc = rowMeans(expr_filt[, is_lusc]) - rowMeans(expr_filt[, is_luad]),
  pval = sapply(seq_len(nrow(expr_filt)), function(i) {
    t.test(expr_filt[i, is_lusc], expr_filt[i, is_luad])$p.value
  })
)
de_results$padj <- p.adjust(de_results$pval, method = "BH")
de_results$neg_log10p <- -log10(de_results$padj)
de_results$direction <- case_when(
  de_results$padj < 0.01 & de_results$log2fc > 1 ~ "Up in LUSC",
  de_results$padj < 0.01 & de_results$log2fc < -1 ~ "Up in LUAD",
  TRUE ~ "NS"
)

# Label top genes
top_de <- de_results |>
  filter(direction != "NS") |>
  arrange(desc(abs(log2fc))) |>
  head(20)

publish(ggplot(de_results, aes(x = log2fc, y = neg_log10p)) +
  geom_point(aes(color = direction), alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("Up in LUSC" = "#c0392b", "Up in LUAD" = "#2166ac",
                                "NS" = "grey60")) +
  geom_text_repel(data = top_de, aes(label = gene),
                  size = 3, max.overlaps = 20) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", color = "grey40") +
  labs(
    title = "Differential expression: LUSC vs LUAD",
    subtitle = "Welch t-test on log2(TPM+1); BH-adjusted p-values",
    x = "log2 fold-change (LUSC - LUAD)",
    y = "-log10(adjusted p-value)"
  ),
  data=de_results)
```

## UMAP

PCA is a linear method — it finds the best flat projection. UMAP (Uniform Manifold
Approximation and Projection) is a non-linear method that preserves local neighborhood
structure, often revealing clusters that PCA compresses together. We use cosine distance,
which is standard for expression data because it focuses on the *direction* of the
expression profile rather than its magnitude.

`n_neighbors = 30` controls the balance between local and global structure (higher values
preserve more global topology). `min_dist = 0.3` prevents points from collapsing into
tight clumps, keeping the embedding readable.

```{r umap}
set.seed(42)
umap_res <- umap(t(expr_filt), n_neighbors = 50, min_dist = 0.3, metric = "cosine")

umap_df <- data.frame(
  UMAP1 = umap_res[, 1],
  UMAP2 = umap_res[, 2],
  stage = clin_tumor$ajcc_pathologic_stage,
  gender = clin_tumor$gender,
  cohort = clin_tumor$cohort,
  vital_status = clin_tumor$vital_status
)
```

```{r umap-cohort}
p <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = cohort)) +
  geom_point(alpha = 0.6, size = 2) +
  labs(title = "UMAP embedding of TCGA lung tumors", subtitle = "Colored by cancer type", color = "Cohort")
publish(p)
```

## K-means clustering

To move from visualization to discrete subtypes, we apply k-means clustering in PCA
space (first 20 components, which capture the majority of variance while filtering out
noise in later components). With LUAD and LUSC combined, we expect at minimum two clusters
reflecting cancer type, with additional clusters capturing within-histology heterogeneity
(e.g. immune-hot vs immune-cold, or proliferative vs differentiated subtypes).

We wrap this in `reproducible()` so that viewers can experiment with the two key
parameters — number of clusters (k) and number of PCs — directly in the GoFigr portal,
without re-running the notebook. The `slider()` calls define interactive widgets; `static()`
passes in the data without making it editable.

```{r kmeans}
reproducible(function(
    k = slider(3, min = 2, max = 6, step = 1),
    n_pcs = slider(20, min = 5, max = 30, step = 5),
    pca_scores = static(pca$x),
    umap_coords = static(umap_df[, c("UMAP1", "UMAP2")])
  ) {
    set.seed(42)
    km <- kmeans(pca_scores[, 1:n_pcs], centers = k, nstart = 25, iter.max = 100)
    plot_df <- data.frame(umap_coords, cluster = factor(km$cluster))

    publish(
      ggplot(plot_df, aes(x = UMAP1, y = UMAP2, color = cluster)) +
        geom_point(alpha = 0.6, size = 2) +
        scale_color_brewer(palette = "Set1") +
        labs(
          title = "Unsupervised clusters on UMAP",
          subtitle = sprintf("k-means (k=%d) on top %d PCs", k, n_pcs),
          color = "Cluster"
        ),
      figure_name = "K-means clusters on UMAP"
    )
  },
  packages = c("ggplot2"),
  name = "K-means clustering",
  interactive=FALSE
)

# Store the default clustering for downstream analyses
set.seed(42)
km <- kmeans(pca$x[, 1:20], centers = 3, nstart = 25, iter.max = 100)
umap_df$cluster <- factor(km$cluster)
clin_tumor$cluster <- factor(km$cluster)
```

### Cluster characterization

Now we ask: what makes each cluster biologically distinct? We compute the mean expression
of every gene within each cluster, then pick the 50 genes with the highest between-cluster
variance — these are the genes that most differentiate the subtypes. The heatmap gives an
at-a-glance view of the molecular identity of each cluster: look for immune genes
concentrated in one cluster (the "hot" subtype), proliferation markers in another, and
lung differentiation markers in the third.

```{r cluster-heatmap}
cluster_means <- sapply(levels(clin_tumor$cluster), function(cl) {
  rowMeans(expr_filt[, clin_tumor$cluster == cl], na.rm = TRUE)
})
colnames(cluster_means) <- paste0("Cluster ", colnames(cluster_means))

# Top variable genes across clusters
cluster_var <- apply(cluster_means, 1, var)
top_genes <- names(sort(cluster_var, decreasing = TRUE))[1:50]

# silent = TRUE suppresses pheatmap's internal plot call, so that only
# gofigR::publish() draws the figure — otherwise it renders twice.
pheatmap(
  cluster_means[top_genes, ],
  scale = "none",
  cluster_cols = FALSE,
  fontsize_row = 8,
  color = colorRampPalette(c("#2166ac", "white", "#b2182b"))(100),
  main = "Top 50 cluster-discriminating genes",
  silent = TRUE
) %>% publish
```

# 2. Immune Landscape

The tumor immune microenvironment is one of the most clinically actionable axes in LUAD.
Immune checkpoint inhibitors (anti-PD-1, anti-PD-L1, anti-CTLA-4) have transformed
treatment for a subset of patients — but response rates are only 20–30%. Understanding
which tumors are "immune-hot" (T-cell infiltrated, checkpoint-expressing) vs.
"immune-cold" (excluded or desert) is key to predicting who will benefit.

## Immune scoring

We define two composite scores using the mean log2(TPM+1) of curated gene sets:

- **Immune activation**: IFNG, GZMB, PRF1, CXCL9, CXCL10, CD8A, CD8B. These are
  hallmarks of cytotoxic T-cell and NK cell activity. This is closely related to the
  "cytolytic activity" score from Rooney et al. (Cell, 2015), which has been validated
  as a predictor of immunotherapy response.
- **Immune suppression**: IDO1, TGFB1, TGFB2, IL10, ARG1, CD274 (PD-L1), VEGFA. These
  represent mechanisms tumors use to evade immune attack — checkpoint ligands, suppressive
  cytokines, and metabolic immune inhibitors.

The `score_genes` helper gracefully handles missing genes (not all ~250 may have mapped
to the TCGA annotation), taking the mean over whichever genes are available.

```{r immune-scores}
score_genes <- function(expr_mat, genes) {
  available <- genes[genes %in% rownames(expr_mat)]
  if (length(available) == 0) return(rep(NA, ncol(expr_mat)))
  colMeans(expr_mat[available, , drop = FALSE], na.rm = TRUE)
}

# Cytolytic / immune activation (Rooney et al., Cell 2015)
activation_genes <- c("IFNG", "GZMB", "PRF1", "CXCL9", "CXCL10", "CD8A", "CD8B")

# Immunosuppressive markers
suppression_genes <- c("IDO1", "TGFB1", "TGFB2", "IL10", "ARG1", "CD274", "VEGFA")

clin_tumor$immune_activation <- score_genes(expr_tumor, activation_genes)
clin_tumor$immune_suppression <- score_genes(expr_tumor, suppression_genes)
```

Plotting activation vs. suppression creates a "landscape" where each tumor occupies a
position. The positive correlation between the two scores is expected — tumors that
attract immune cells also upregulate suppressive checkpoints as an adaptive resistance
mechanism. The clinically interesting samples are those with high activation but low
suppression (potentially responsive to immunotherapy without additional checkpoint
blockade), or high suppression with low activation (deeply suppressed microenvironment).

```{r hot-cold-scatter}
publish({ggplot(clin_tumor, aes(x = immune_activation, y = immune_suppression)) +
  geom_point(aes(color = cluster), alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", color = "grey30", linetype = "dashed", se = FALSE) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Immune activation vs. suppression",
    subtitle = "Each point is a tumor sample; color = cluster",
    x = "Immune activation score\n(IFNG, GZMB, PRF1, CXCL9, CXCL10, CD8A/B)",
    y = "Immune suppression score\n(IDO1, TGFB1/2, IL10, ARG1, PD-L1, VEGFA)",
    color = "Cluster"
  )})
```

## Checkpoint expression by cluster

Here we ask which immune checkpoints are most expressed in each transcriptomic cluster.
This has direct clinical implications: if a cluster co-expresses PD-L1 and LAG-3, that
suggests dual checkpoint blockade might be warranted. We use common gene symbols and map
them to their clinical names (e.g. CD274 = PD-L1, HAVCR2 = TIM-3) for readability.

```{r checkpoint-heatmap}
checkpoints <- c("CD274", "PDCD1", "CTLA4", "LAG3", "HAVCR2", "TIGIT", "IDO1")
checkpoints_avail <- checkpoints[checkpoints %in% rownames(expr_tumor)]

checkpoint_by_cluster <- sapply(levels(clin_tumor$cluster), function(cl) {
  rowMeans(expr_tumor[checkpoints_avail, clin_tumor$cluster == cl, drop = FALSE], na.rm = TRUE)
})
colnames(checkpoint_by_cluster) <- paste0("Cluster ", colnames(checkpoint_by_cluster))

# Rename for readability
display_names <- c(
  CD274 = "PD-L1", PDCD1 = "PD-1", CTLA4 = "CTLA-4",
  LAG3 = "LAG-3", HAVCR2 = "TIM-3", TIGIT = "TIGIT", IDO1 = "IDO1"
)
rownames(checkpoint_by_cluster) <- display_names[rownames(checkpoint_by_cluster)]

pheatmap(
  checkpoint_by_cluster,
  scale = "row",
  cluster_cols = FALSE,
  color = colorRampPalette(c("#2166ac", "white", "#b2182b"))(100),
  main = "Immune checkpoint expression by cluster",
  silent = TRUE
) %>% publish(figure_name = "Immune checkpoint expression by cluster")
```

## Checkpoint correlations with immune activation

Not all checkpoints are created equal. Some (like PD-1 and CTLA-4) are markers of
activated T cells and correlate strongly with immune infiltration. Others (like IDO1)
represent a tumor-intrinsic suppressive program. This bar chart shows Pearson
correlation of each checkpoint with our immune activation score — highly correlated
checkpoints are the ones most likely to be co-opted by immune-infiltrated tumors as an
adaptive resistance strategy.

```{r checkpoint-cor}
checkpoint_expr <- t(expr_tumor[checkpoints_avail, , drop = FALSE])
colnames(checkpoint_expr) <- display_names[colnames(checkpoint_expr)]

cor_df <- data.frame(
  checkpoint = colnames(checkpoint_expr),
  correlation = sapply(seq_len(ncol(checkpoint_expr)), function(i) {
    cor(checkpoint_expr[, i], clin_tumor$immune_activation, use = "complete.obs")
  })
) |> mutate(checkpoint = reorder(checkpoint, correlation))

publish({ggplot(cor_df, aes(x = correlation, y = checkpoint)) +
  geom_col(aes(fill = correlation), show.legend = FALSE) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Checkpoint correlation with immune activation",
    subtitle = "Pearson r across tumor samples",
    x = "Correlation with immune activation score",
    y = NULL
  )})
```

# 3. Oncogenic Pathway Crosstalk

Cancer is not driven by single genes in isolation — it's the interplay between signaling
pathways that determines tumor phenotype and treatment vulnerability. Here we score each
sample across 8 major pathway categories using mean log2(TPM+1) expression, then ask
which pathways tend to co-activate (positive correlation) or are mutually exclusive
(negative correlation) across tumors.

## Pathway scoring

For each pathway, we compute a composite score as the mean log2(TPM+1) expression of its
constituent genes. This is a simple but robust approach: it's less noisy than any single
gene, it's interpretable, and it correlates well with more sophisticated methods like
GSVA or ssGSEA for well-curated gene sets of this size.

```{r pathway-scores}
pathway_lists <- list(
  `RTK/RAS/MAPK` = c("EGFR", "KRAS", "BRAF", "ERBB2", "MET", "ALK", "ROS1",
                      "NF1", "MAPK1", "MAPK3", "SOS1", "GRB2"),
  `PI3K/AKT/mTOR` = c("PIK3CA", "PIK3CB", "PIK3R1", "AKT1", "AKT2", "MTOR",
                       "PTEN", "TSC1", "TSC2", "RICTOR", "RPTOR"),
  `Wnt` = c("CTNNB1", "APC", "GSK3B", "AXIN1", "LEF1", "TCF7L2",
            "WNT3A", "WNT5A", "WNT7A", "FZD7"),
  `DDR/Cell Cycle` = c("ATM", "ATR", "CHEK1", "CHEK2", "PARP1", "CDK1", "CDK2",
                        "CCNA2", "CCNB1", "PLK1", "AURKA", "E2F1", "FOXM1"),
  `Immune` = c("IFNG", "GZMB", "PRF1", "CD8A", "CD274", "PDCD1",
               "CTLA4", "CXCL9", "CXCL10", "TNF"),
  `EMT` = c("CDH1", "CDH2", "VIM", "SNAI1", "SNAI2", "ZEB1", "ZEB2",
            "TWIST1", "FN1", "MMP2", "MMP9"),
  `Apoptosis` = c("BAX", "BAK1", "BCL2", "BCL2L1", "MCL1", "CASP3",
                  "CASP8", "CASP9", "BIRC5", "XIAP"),
  `Metabolism` = c("HK2", "PKM", "LDHA", "SLC2A1", "FASN", "ACLY",
                   "GLS", "SLC7A11", "GPX4", "NFE2L2")
)

pathway_scores <- sapply(pathway_lists, function(genes) {
  score_genes(expr_filt, genes)
})

clin_tumor <- cbind(clin_tumor, as.data.frame(pathway_scores, check.names = FALSE))
```

## Pathway correlation heatmap

The correlation matrix reveals the dependencies between oncogenic programs. Key patterns
to look for:

- **DDR/Cell Cycle and Metabolism**: often co-activated in aggressive, proliferative
  tumors that upregulate glycolysis (the Warburg effect) to fuel rapid division.
- **Immune and EMT**: the relationship here previews our analysis in Section 4. EMT
  (epithelial-to-mesenchymal transition) is associated with immune exclusion.
- **RTK/RAS/MAPK and PI3K/AKT**: these are parallel signaling cascades downstream of
  receptor tyrosine kinases. Co-activation suggests redundancy that may explain
  resistance to single-pathway inhibitors.

```{r pathway-cor-heatmap}
pathway_cor <- cor(pathway_scores, use = "pairwise.complete.obs")

pheatmap(
  pathway_cor,
  color = colorRampPalette(c("#2166ac", "white", "#b2182b"))(100),
  breaks = seq(-1, 1, length.out = 101),
  display_numbers = TRUE,
  number_format = "%.2f",
  fontsize_number = 9,
  main = "Pathway co-activation across lung tumors (LUAD + LUSC)",
  silent = TRUE
) %>% publish
```

## KRAS/STK11/KEAP1 axis

The co-occurrence of KRAS activation with STK11 and KEAP1 loss defines the most
treatment-resistant LUAD subgroup, characterized by immune exclusion and
metabolic rewiring (Skoulidis et al., Cancer Discovery 2015). This axis is specific to
LUAD biology — KRAS mutations are rare in LUSC — so we restrict this analysis to LUAD
samples only.

The median split for "KRAS-high" and "STK11-low" is a common but arbitrary choice. We
use `reproducible()` with percentile sliders so viewers can test how sensitive the immune
exclusion finding is to different cutoff definitions.

```{r kras-stk11-keap1}
# Restrict to LUAD — KRAS/STK11 axis is LUAD-specific
luad_idx <- clin_tumor$cohort == "LUAD"
expr_luad <- expr_tumor[, luad_idx]
clin_luad <- clin_tumor[luad_idx, ]

axis_genes <- c("KRAS", "STK11", "KEAP1")
axis_avail <- axis_genes[axis_genes %in% rownames(expr_luad)]

axis_df <- data.frame(
  t(expr_luad[axis_avail, , drop = FALSE]),
  immune_activation = clin_luad$immune_activation,
  cluster = clin_luad$cluster,
  check.names = FALSE
)

if (all(c("KRAS", "STK11") %in% axis_avail)) {
  reproducible(function(
      kras_percentile = slider(50, min = 25, max = 75, step = 5),
      stk11_percentile = slider(50, min = 25, max = 75, step = 5),
      data = static(axis_df)
    ) {
      kras_thresh <- quantile(data$KRAS, kras_percentile / 100, na.rm = TRUE)
      stk11_thresh <- quantile(data$STK11, stk11_percentile / 100, na.rm = TRUE)

      data$subgroup <- case_when(
        data$KRAS > kras_thresh & data$STK11 < stk11_thresh ~ "KRAS-high\nSTK11-low",
        data$KRAS > kras_thresh ~ "KRAS-high\nSTK11-intact",
        TRUE ~ "KRAS-low"
      )

      publish(
        ggplot(data, aes(x = subgroup, y = immune_activation, fill = subgroup)) +
          geom_boxplot(outlier.alpha = 0.4) +
          geom_jitter(width = 0.2, alpha = 0.2, size = 1) +
          scale_fill_brewer(palette = "Set2") +
          labs(
            title = "Immune activation by KRAS/STK11 subgroup (LUAD only)",
            subtitle = sprintf("KRAS > %dth %%ile, STK11 < %dth %%ile",
                               kras_percentile, stk11_percentile),
            x = NULL,
            y = "Immune activation score"
          ) +
          theme(legend.position = "none"),
        figure_name = "KRAS/STK11 immune exclusion"
      )
    },
    packages = c("ggplot2", "dplyr"),
    name = "KRAS/STK11 subgroup analysis"
  )
}
```

# 4. EMT-Immune Exclusion Axis

Epithelial-to-mesenchymal transition (EMT) is a developmental program that cancer cells
co-opt to become more invasive and migratory. A growing body of evidence shows that
mesenchymal tumors also tend to exclude cytotoxic T cells from the tumor bed (Thompson
et al., Cancer Immunol Res 2020; Dongre & Weinberg, Nat Rev Mol Cell Biol 2019). This
has major implications for immunotherapy: if EMT drives immune exclusion, then reversing
EMT (or targeting the mesenchymal state) could sensitize tumors to checkpoint blockade.

We construct an EMT score as the difference between mesenchymal marker expression (CDH2/
N-cadherin, VIM/Vimentin, ZEB1, ZEB2, SNAI1/Snail, SNAI2/Slug, FN1/Fibronectin) and
epithelial marker expression (CDH1/E-cadherin). High values indicate a mesenchymal
phenotype; low values indicate an epithelial phenotype.

Crucially, we must analyze LUAD and LUSC separately here. LUSC is a squamous carcinoma —
inherently more mesenchymal with high keratins and distinct differentiation programs. Pooling
the two would produce a spurious EMT-immune correlation driven entirely by histological
differences rather than within-tumor biology.

For the immune side, we use a focused CD8+ T-cell cytotoxicity score: CD8A, CD8B (T-cell
co-receptor), GZMB (granzyme B, the effector protease), and PRF1 (perforin, the pore-
forming protein). Together these capture whether cytotoxic T cells are not just present
but actively armed.

```{r emt-score}
mesenchymal <- c("CDH2", "VIM", "ZEB1", "ZEB2", "SNAI1", "SNAI2", "FN1")
epithelial <- c("CDH1")

clin_tumor$emt_score <- score_genes(expr_tumor, mesenchymal) -
                        score_genes(expr_tumor, epithelial)

cd8_genes <- c("CD8A", "CD8B", "GZMB", "PRF1")
clin_tumor$cd8_score <- score_genes(expr_tumor, cd8_genes)

clin_tumor$stage_simple <- case_when(
  grepl("Stage I[^IV]|Stage IA|Stage IB|^Stage I$", clin_tumor$ajcc_pathologic_stage) ~ "I",
  grepl("Stage II", clin_tumor$ajcc_pathologic_stage) ~ "II",
  grepl("Stage III", clin_tumor$ajcc_pathologic_stage) ~ "III",
  grepl("Stage IV", clin_tumor$ajcc_pathologic_stage) ~ "IV",
  TRUE ~ NA_character_
)
```

```{r emt-immune-scatter}
# Compute per-cohort correlations for subtitles
emt_cor <- clin_tumor |>
  group_by(cohort) |>
  summarise(
    r = cor(emt_score, cd8_score, use = "complete.obs"),
    p = cor.test(emt_score, cd8_score)$p.value,
    .groups = "drop"
  ) |>
  mutate(label = sprintf("r = %.2f (p = %.2g)", r, p))

publish(ggplot(clin_tumor, aes(x = emt_score, y = cd8_score)) +
  geom_point(aes(color = cluster), alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", color = "grey20", se = TRUE) +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~cohort) +
  geom_text(data = emt_cor, aes(label = label),
            x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5, size = 3.5,
            inherit.aes = FALSE) +
  labs(
    title = "EMT score vs. CD8+ T-cell infiltration",
    subtitle = "Faceted by cancer type to avoid confounding by histology",
    x = "EMT score (mesenchymal - epithelial)",
    y = "CD8+ T-cell score (CD8A/B, GZMB, PRF1)",
    color = "Cluster"
  ))
```

# 5. Survival Analysis

Everything so far has been molecular characterization — interesting biology, but the
"so what?" for a clinical audience is whether any of it predicts patient outcomes. TCGA
provides overall survival (OS) data: `days_to_death` for deceased patients and
`days_to_last_follow_up` for censored (still alive at last contact) patients.

We focus on two of the most robustly validated prognostic signatures in NSCLC:

1. **Proliferation score** — a mitotic gene index based on cell-cycle regulators. This
   is consistently the strongest transcriptomic predictor of survival across NSCLC subtypes
   (Gentles et al., Nat Med 2015). High proliferation = poor prognosis.
2. **NKX2-1/TRU differentiation** (LUAD only) — NKX2-1 (TTF-1) expression marks the
   Terminal Respiratory Unit (TRU) subtype identified in the TCGA LUAD landmark paper
   (Cancer Genome Atlas Research Network, Nature 2014). TRU tumors retain alveolar
   differentiation and have significantly better outcomes. NKX2-1 is the single most
   validated prognostic gene in LUAD (Barletta et al., JCO 2009).

Since LUAD and LUSC have different baseline prognoses and biology, we analyze each
cancer type separately.

## Prepare survival data

We combine `days_to_death` and `days_to_last_follow_up` into a single overall survival
time, and code the event indicator from `vital_status`. Samples missing survival data
or with non-positive follow-up are excluded.

```{r survival-prep}
clin_tumor$os_time <- coalesce(
  as.numeric(clin_tumor$days_to_death),
  as.numeric(clin_tumor$days_to_last_follow_up)
)
clin_tumor$os_event <- as.integer(clin_tumor$vital_status == "Dead")
clin_tumor$sample_id <- colnames(expr_tumor)

# Filter to samples with valid survival data
surv_df <- clin_tumor |>
  filter(!is.na(os_time), os_time > 0, !is.na(os_event))

cat(sprintf("Samples with survival data: %d (events: %d)\n",
            nrow(surv_df), sum(surv_df$os_event)))
```

## Proliferation score

The proliferation (mitotic index) score is the most robust transcriptomic prognostic
marker in solid tumors. We compute it as the mean log2(TPM+1) of core cell-cycle and
mitotic genes — these capture the tumor's intrinsic proliferative rate and, indirectly,
TP53 functional status (Whitfield et al., Mol Biol Cell 2002; Gentles et al., Nat Med
2015).

```{r proliferation-score}
prolif_genes <- c("MKI67", "TOP2A", "CCNB1", "CCNA2", "CDK1", "CDK2", "AURKA",
                  "AURKB", "PLK1", "BUB1", "TTK", "CDC20", "FOXM1", "E2F1", "BIRC5")

surv_df$proliferation <- score_genes(expr_tumor[, surv_df$sample_id], prolif_genes)

cat(sprintf("Proliferation score uses %d / %d genes\n",
            sum(prolif_genes %in% rownames(expr_tumor)), length(prolif_genes)))
```

We split at the cohort-specific median to produce Kaplan-Meier curves. The expected
result: high-proliferation tumors have significantly worse OS in both LUAD and LUSC.

```{r km-proliferation}
surv_df <- surv_df |>
  group_by(cohort) |>
  mutate(prolif_group = ifelse(
    proliferation > median(proliferation, na.rm = TRUE),
    "High", "Low"
  )) |>
  ungroup()

for (coh in c("LUAD", "LUSC")) {
  df <- surv_df |> filter(cohort == coh)
  fit <- survfit(Surv(os_time, os_event) ~ prolif_group, data = df)
  ggsurvplot(
    fit, data = df, pval = TRUE, risk.table = TRUE,
    palette = c("#c0392b", "#2980b9"),
    title = sprintf("Overall survival by proliferation score (%s)", coh),
    xlab = "Days", ylab = "Survival probability",
    legend.title = "Proliferation",
    risk.table.height = 0.25,
    ggtheme = theme_minimal(base_size = 14)
  ) %>% publish
}
```

## NKX2-1 (TTF-1) expression in LUAD

NKX2-1 is the master transcription factor of alveolar differentiation. In LUAD, high
NKX2-1 expression marks the TRU subtype — well-differentiated tumors that retain alveolar
identity and have the best prognosis. Loss of NKX2-1 is associated with de-differentiation,
invasiveness, and poor outcomes (Winslow et al., Nature 2011).

This is LUAD-specific: NKX2-1 is silenced in LUSC (squamous tumors use TP63/SOX2 for
lineage identity instead).

```{r km-nkx2-1}
# NKX2-1 may be stored under its gene symbol or alias
nkx_gene <- intersect(c("NKX2-1", "TTF1"), rownames(expr_tumor))

if (length(nkx_gene) > 0) {
  surv_df$nkx2_1 <- as.numeric(expr_tumor[nkx_gene[1], surv_df$sample_id])

  df_luad <- surv_df |> filter(cohort == "LUAD")
  df_luad$nkx_group <- ifelse(
    df_luad$nkx2_1 > median(df_luad$nkx2_1, na.rm = TRUE),
    "NKX2-1 High", "NKX2-1 Low"
  )

  fit_nkx <- survfit(Surv(os_time, os_event) ~ nkx_group, data = df_luad)
  ggsurvplot(
    fit_nkx, data = df_luad, pval = TRUE, risk.table = TRUE,
    palette = c("#27ae60", "#8e44ad"),
    title = "Overall survival by NKX2-1 expression (LUAD only)",
    subtitle = "NKX2-1 marks TRU subtype — high = alveolar differentiation, better prognosis",
    xlab = "Days", ylab = "Survival probability",
    legend.title = "NKX2-1",
    risk.table.height = 0.25,
    ggtheme = theme_minimal(base_size = 14)
  ) %>% publish
} else {
  cat("NKX2-1 not found in expression matrix\n")
}
```

## EMT score and survival

EMT-high tumors tend to be more invasive and to exclude cytotoxic T cells (Section 4).
We test whether this translates to worse survival within each cancer type.

```{r km-emt}
surv_df <- surv_df |>
  group_by(cohort) |>
  mutate(emt_group = ifelse(
    emt_score > median(emt_score, na.rm = TRUE),
    "High (mesenchymal)", "Low (epithelial)"
  )) |>
  ungroup()

for (coh in c("LUAD", "LUSC")) {
  df <- surv_df |> filter(cohort == coh)
  fit <- survfit(Surv(os_time, os_event) ~ emt_group, data = df)
  ggsurvplot(
    fit, data = df, pval = TRUE, risk.table = TRUE,
    palette = c("#8e44ad", "#27ae60"),
    title = sprintf("Overall survival by EMT score (%s)", coh),
    xlab = "Days", ylab = "Survival probability",
    legend.title = "EMT",
    risk.table.height = 0.25,
    ggtheme = theme_minimal(base_size = 14)
  ) %>% publish
}
```

## Multivariate Cox regression

The Kaplan-Meier analyses above are univariate — they don't account for confounders.
Stage and histology are strong clinical prognostic factors, so we need to ask: do the
molecular scores add information *beyond* what stage and cancer type already tell us?

A Cox proportional hazards model lets us estimate the hazard ratio (HR) for each
covariate while adjusting for all others. An HR > 1 means higher values of that feature
are associated with increased risk of death; HR < 1 means they're protective. The forest
plot makes it easy to see which features cross the HR = 1 line (non-significant) vs.
which are clearly displaced from it.

```{r cox}
surv_df$stage_numeric <- as.numeric(factor(
  surv_df$stage_simple,
  levels = c("I", "II", "III", "IV")
))
surv_df$is_lusc <- as.integer(surv_df$cohort == "LUSC")

cox_fit <- coxph(
  Surv(os_time, os_event) ~
    stage_numeric + is_lusc + proliferation + emt_score +
    immune_activation,
  data = surv_df
)

cox_summary <- broom::tidy(cox_fit, conf.int = TRUE, exponentiate = TRUE)

# Readable labels
label_map <- c(
  stage_numeric = "Stage (I-IV)",
  is_lusc = "LUSC (vs LUAD)",
  proliferation = "Proliferation",
  emt_score = "EMT score",
  immune_activation = "Immune activation"
)
cox_summary$label <- label_map[cox_summary$term]

publish(
  ggplot(cox_summary, aes(x = estimate, y = reorder(label, estimate))) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  labs(
    title = "Multivariate Cox regression: hazard ratios",
    subtitle = "Adjusted for stage, histology, and molecular scores",
    x = "Hazard ratio (95% CI)",
    y = NULL
  ) +
  scale_x_log10())
```

# Session Info

```{r session-info}
sessionInfo()
```
