---
title: "TCGA Lung Cancer: LUAD vs LUSC"
author: "GoFigr Webinar"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    fig-width: 10
    fig-height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  dpi = 150
)
```

# Setup

```{r libraries}
library(gofigR)
library(SummarizedExperiment)
library(ggplot2)
library(dplyr)
library(uwot)
library(survival)
library(survminer)
library(ggrepel)
```

## Enabling GoFigr

[GoFigr](https://gofigr.io) automatically captures, versions, and indexes every figure
you produce. Once enabled, every plot is published — no extra calls needed. Browse and
share your figures at [app.gofigr.io](https://app.gofigr.io).

```{r gofigr}
gofigR::enable(
  analysis_name = "TCGA Lung Cancer Analysis (LUAD + LUSC)"
)
```

# Data Loading

We work with two TCGA lung cancer cohorts: **LUAD** (Lung Adenocarcinoma) and **LUSC**
(Lung Squamous Cell Carcinoma) — the two most common subtypes of non-small cell lung
cancer (NSCLC). The data contains RNA-Seq TPM values for ~1200 curated cancer and
highly-expressed genes across ~1000 tumor samples.

Notice how we use `gofigR::sync_file` to (1) store the file as an "Asset" in the GoFigr portal,
and (2) link this specific file to every figure generated here.

If that file ever changes, we will upload a new revision automatically. Your linkage from 
figures to specific file versions remains bulletproof -- old figures still link to the
specific older version that generated them.

```{r load-data}
data_dir <- file.path(getwd(), "data")
rds_path <- file.path(data_dir, "TCGA-LUNG_TPM_SE_subset.rds")

if (!file.exists(rds_path)) {
  dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
  message("Downloading TCGA data...")
  download.file(
    "https://cdn.gofigr.io/data/TCGA-LUNG_TPM_SE_subset.rds",
    rds_path,
    mode = "wb"
  )
}

# GoFigr can track and version your data for you!
gofigR::sync_file(rds_path)

se <- readRDS(rds_path)

# Log-transform and map to gene symbols
expr_raw <- assay(se, "tpm_unstrand")
expr <- log2(expr_raw + 1)
rownames(expr) <- make.unique(rowData(se)$gene_name)

clinical <- as.data.frame(colData(se))

# Restrict to primary tumors
tumor_idx <- clinical$sample_type == "Primary Tumor"
expr_tumor <- expr[, tumor_idx]
clin_tumor <- clinical[tumor_idx, ]

cat(sprintf("Tumor samples: %d (%d LUAD, %d LUSC) | Genes: %d\n",
            ncol(expr_tumor),
            sum(clin_tumor$cohort == "LUAD"),
            sum(clin_tumor$cohort == "LUSC"),
            nrow(expr_tumor)))
```

# LUAD vs LUSC: Transcriptomic Differences

## PCA

PCA projects the high-dimensional expression data onto axes of maximum variance. With
two distinct cancer types, we expect the dominant axis to separate LUAD from LUSC.

```{r pca}
gene_var <- apply(expr_tumor, 1, var, na.rm = TRUE)
expr_filt <- expr_tumor[gene_var > 0, ]

pca <- prcomp(t(expr_filt), center = TRUE, scale. = FALSE)
pct_var <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

pca_df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  cohort = clin_tumor$cohort
)

# Keep the top 30 PCs for downstream clustering
n_pcs_max <- min(30, ncol(pca$x))
pca_scores_df <- as.data.frame(pca$x[, seq_len(n_pcs_max)])

p <- ggplot(pca_df, aes(x = PC1, y = PC2, color = cohort)) +
  geom_point(alpha = 0.6, size = 2) +
  labs(
    x = sprintf("PC1 (%.1f%%)", pct_var[1]),
    y = sprintf("PC2 (%.1f%%)", pct_var[2]),
    title = "PCA of TCGA lung tumors (LUAD + LUSC)",
    color = "Cohort"
  )
p %>% publish
```

## Top PCA loadings

The top PC1 loadings reveal what drives the separation: squamous markers (keratins, TP63,
SOX2) vs adenocarcinoma markers (NKX2-1/TTF-1, surfactant proteins, NAPSA).

```{r pca-loadings}
loadings <- as.data.frame(pca$rotation[, 1:3])
loadings$gene <- rownames(loadings)

top_pc1 <- loadings |>
  arrange(desc(abs(PC1))) |>
  head(15) |>
  mutate(gene = reorder(gene, abs(PC1)))

publish({ggplot(top_pc1, aes(x = PC1, y = gene)) +
  geom_col(aes(fill = PC1 > 0), show.legend = FALSE) +
  scale_fill_manual(values = c("TRUE" = "#c0392b", "FALSE" = "#2980b9")) +
  labs(
    title = "Top 15 genes driving PC1",
    x = "Loading on PC1",
    y = NULL
  )})
```

## Volcano plot

We directly quantify expression differences with a Welch t-test per gene.

```{r de-luad-lusc}
is_luad <- clin_tumor$cohort == "LUAD"
is_lusc <- clin_tumor$cohort == "LUSC"

de_results <- data.frame(
  gene = rownames(expr_filt),
  log2fc = rowMeans(expr_filt[, is_lusc]) - rowMeans(expr_filt[, is_luad]),
  pval = sapply(seq_len(nrow(expr_filt)), function(i) {
    t.test(expr_filt[i, is_lusc], expr_filt[i, is_luad])$p.value
  })
)
de_results$padj <- p.adjust(de_results$pval, method = "BH")
de_results$neg_log10p <- -log10(de_results$padj)
de_results$direction <- case_when(
  de_results$padj < 0.01 & de_results$log2fc > 1 ~ "Up in LUSC",
  de_results$padj < 0.01 & de_results$log2fc < -1 ~ "Up in LUAD",
  TRUE ~ "NS"
)

top_de <- de_results |>
  filter(direction != "NS") |>
  arrange(desc(abs(log2fc))) |>
  head(20)

publish(ggplot(de_results, aes(x = log2fc, y = neg_log10p)) +
  geom_point(aes(color = direction), alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c("Up in LUSC" = "#c0392b", "Up in LUAD" = "#2166ac",
                                "NS" = "grey60")) +
  geom_text_repel(data = top_de, aes(label = gene),
                  size = 3, max.overlaps = 20) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", color = "grey40") +
  labs(
    title = "Differential expression: LUSC vs LUAD",
    subtitle = "Welch t-test on log2(TPM+1); BH-adjusted p-values",
    x = "log2 fold-change (LUSC - LUAD)",
    y = "-log10(adjusted p-value)"
  ),
  data=de_results)
```

## UMAP

UMAP preserves local neighborhood structure, often revealing clusters that PCA compresses.

```{r umap}
set.seed(42)
umap_res <- umap(t(expr_filt), n_neighbors = 50, min_dist = 0.3, metric = "cosine")

umap_df <- data.frame(
  UMAP1 = umap_res[, 1],
  UMAP2 = umap_res[, 2],
  cohort = clin_tumor$cohort
)

p <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = cohort)) +
  geom_point(alpha = 0.6, size = 2) +
  labs(title = "UMAP embedding of TCGA lung tumors", color = "Cohort")
publish(p)
```

## K-means clustering

We apply k-means in PCA space. This is wrapped in `reproducible()` — a GoFigr feature
that lets viewers adjust k and n_pcs interactively in the portal without re-running the
notebook.

```{r kmeans}
reproducible(function(
    k = slider(3, min = 2, max = 6, step = 1),
    n_pcs = slider(20, min = 5, max = 30, step = 5),
    pca_scores = static(pca_scores_df),
    umap_coords = static(umap_df[, c("UMAP1", "UMAP2")])
  ) {
    set.seed(42)
    km <- kmeans(as.matrix(pca_scores[, 1:n_pcs]), centers = k, nstart = 25, iter.max = 100)
    plot_df <- data.frame(umap_coords, cluster = factor(km$cluster))

    publish(
      ggplot(plot_df, aes(x = UMAP1, y = UMAP2, color = cluster)) +
        geom_point(alpha = 0.6, size = 2) +
        scale_color_brewer(palette = "Set1") +
        labs(
          title = "Unsupervised clusters on UMAP",
          subtitle = sprintf("k-means (k=%d) on top %d PCs", k, n_pcs),
          color = "Cluster"
        ),
      figure_name = "K-means clusters on UMAP"
    )
  },
  packages = c("ggplot2", "stats"),
  name = "K-means clustering",
  interactive=FALSE # set to TRUE to show a shiny gadget. It's very experimental, but please try it and report issues!
)
```

# Survival Analysis

Do these molecular differences predict patient outcomes? We test two well-validated
prognostic signatures, wrapped in `reproducible()` so viewers can re-stratify by
clinical covariates (cohort, gender, smoking status, stage) directly in the GoFigr
portal.

```{r survival-prep}
score_genes <- function(expr_mat, genes) {
  available <- genes[genes %in% rownames(expr_mat)]
  if (length(available) == 0) return(rep(NA, ncol(expr_mat)))
  colMeans(expr_mat[available, , drop = FALSE], na.rm = TRUE)
}

prolif_genes <- c("MKI67", "TOP2A", "CCNB1", "CCNA2", "CDK1", "CDK2", "AURKA",
                  "AURKB", "PLK1", "BUB1", "TTK", "CDC20", "FOXM1", "E2F1", "BIRC5")

clin_tumor$os_time <- coalesce(
  as.numeric(clin_tumor$days_to_death),
  as.numeric(clin_tumor$days_to_last_follow_up)
)
clin_tumor$os_event <- as.integer(clin_tumor$vital_status == "Dead")
clin_tumor$sample_id <- colnames(expr_tumor)
clin_tumor$proliferation <- score_genes(expr_tumor, prolif_genes)

# Simplify clinical covariates
clin_tumor$stage_simple <- case_when(
  grepl("Stage I[^IV]|Stage IA|Stage IB|^Stage I$", clin_tumor$ajcc_pathologic_stage) ~ "I",
  grepl("Stage II", clin_tumor$ajcc_pathologic_stage) ~ "II",
  grepl("Stage III", clin_tumor$ajcc_pathologic_stage) ~ "III",
  grepl("Stage IV", clin_tumor$ajcc_pathologic_stage) ~ "IV",
  TRUE ~ NA_character_
)

clin_tumor$smoking <- case_when(
  clin_tumor$tobacco_smoking_status == "Lifelong Non-Smoker" ~ "Never",
  grepl("Current Smoker", clin_tumor$tobacco_smoking_status) ~ "Current",
  grepl("Reformed", clin_tumor$tobacco_smoking_status) ~ "Former",
  TRUE ~ NA_character_
)

clin_tumor$age_group <- ifelse(
  as.numeric(clin_tumor$age_at_diagnosis) / 365.25 >= 67,
  ">=67", "<67"
)

# Build a clean survival data frame with only the columns we need —
# the full colData has list/raw columns that can't be serialized by reproducible()
surv_df <- data.frame(
  sample_id = clin_tumor$sample_id,
  cohort = clin_tumor$cohort,
  gender = clin_tumor$gender,
  stage_simple = clin_tumor$stage_simple,
  smoking = clin_tumor$smoking,
  os_time = clin_tumor$os_time,
  os_event = clin_tumor$os_event,
  proliferation = clin_tumor$proliferation,
  stringsAsFactors = FALSE
) |>
  filter(!is.na(os_time), os_time > 0, !is.na(os_event))

cat(sprintf("Samples with survival data: %d (events: %d)\n",
            nrow(surv_df), sum(surv_df$os_event)))
```

## Proliferation score

The proliferation score is a robust transcriptomic prognostic
marker in solid tumors (Gentles et al., Nat Med 2015). High proliferation = poor
prognosis. Use the dropdowns to explore how the effect varies by clinical subgroup.

```{r km-proliferation}
reproducible(function(
    cohort = dropdown("LUAD", choices = c("LUAD", "LUSC", "Both")),
    gender = dropdown("All", choices = c("All", "male", "female")),
    smoking = dropdown("All", choices = c("All", "Never", "Former", "Current")),
    stage = dropdown("All", choices = c("All", "I", "II", "III", "IV")),
    data = static(surv_df)
  ) {
    df <- data
    if (cohort != "Both") df <- df[df$cohort == cohort, ]
    if (gender != "All") df <- df[df$gender == gender, ]
    if (smoking != "All") df <- df[!is.na(df$smoking) & df$smoking == smoking, ]
    if (stage != "All") df <- df[!is.na(df$stage_simple) & df$stage_simple == stage, ]

    df$group <- ifelse(
      df$proliferation > median(df$proliferation, na.rm = TRUE),
      "High", "Low"
    )

    subtitle <- paste(c(
      cohort,
      if (gender != "All") gender,
      if (smoking != "All") paste0(smoking, " smoker"),
      if (stage != "All") paste0("Stage ", stage)
    ), collapse = " | ")

    fit <- survfit(Surv(os_time, os_event) ~ group, data = df)
    publish(
      ggsurvplot(
        fit, data = df, pval = TRUE, risk.table = FALSE,
        palette = c("#c0392b", "#2980b9"),
        title = "Overall survival by proliferation score",
        subtitle = sprintf("%s (n=%d)", subtitle, nrow(df)),
        xlab = "Days", ylab = "Survival probability",
        legend.title = "Proliferation",
        risk.table.height = 0.25,
        ggtheme = theme_minimal(base_size = 14)
      ),
      figure_name = "Survival by proliferation"
    )
  },
  packages = c("survival", "survminer", "ggplot2", "stats"),
  name = "Proliferation survival",
)
```

# Session Info

```{r session-info}
sessionInfo()
```
